---
title: "fars_analysis"
author: "Rebecca Foos"
date: "November 1, 2017"
output: word_document
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(tidyr)
library(dplyr)
library(tibble)
library(ggplot2)
library(ggthemes)
library(knitr)
library(purrr)

load("../data/clean_fars.RData")
```

##FARS Results 

###Percentages of drivers who testes positive for drugs by drug type, sex, and year strata

```{r table_1, fig.width = 8}
clean_fars %>% 
  mutate(year_cat = cut(year, breaks = c(1999, 2002, 2006, 2010),
                        labels = c("1999-2002","2003-2006",
                                   "2007-2010"),
                        include.lowest = TRUE, right = TRUE)) %>% 
  filter(!is.na(sex)) %>% 
  group_by(drug_type, sex, year_cat) %>% 
  summarize(n_non_missing = sum(!is.na(positive_for_drug)),
            positive_test = sum(positive_for_drug, na.rm = TRUE),
            perc_positive = round(100 * positive_test/n_non_missing, 1)) %>% 
  select(drug_type, sex, year_cat, perc_positive) %>% 
  unite(sex_year_cat, sex, year_cat) %>% 
  spread(sex_year_cat, perc_positive) %>% 
  knitr::kable(col.names = c("Drug Type", "F 1999-2002",
                             "F 2003-2006", "F 2007-2010",
                             "M 1999-2002", "M 2003-2006",
                             "M 2007-2010"))
```

###Fig. 1: Prevalence of non-alcohol drugs in driver fatalities by year and age strata

```{r figure1, fig.width = 8}
fig1 <- clean_fars %>% 
  filter(drug_type != "Alcohol") %>%
  filter(!is.na(agecat)) %>%
  group_by(unique_id, year, agecat) %>% 
  summarize(positive_tests = sum(positive_for_drug, na.rm = TRUE), positive = any(positive_tests > 0)) %>% 
  ungroup() %>% 
  group_by(year, agecat) %>% 
  summarize(total_tests = length(positive), positive = sum(positive), percent_positive = round(100 * positive/total_tests,1))

plot_f1 <- fig1 %>%
  ggplot(aes(x = year, y = percent_positive, position = agecat))+
  geom_point(aes(shape = agecat))+
  geom_line()+
  theme_few()+
  labs(x = "Year", y = "Positive for Non-alcohol Drugs, %", shape = "Age")+
  scale_y_continuous(limits = c(0, 35))
plot_f1
```

###Fig. 2: Prevalence of non-alcohol drugs in driver fatalities by year and drug type

```{r figure2, fig.width = 8}
fig2 <- clean_fars %>% 
  filter(drug_type != "Alcohol") %>% 
  filter(!is.na(positive_for_drug)) %>% 
  group_by(drug_type, year) %>% 
  summarize(n_not_NA = sum(!is.na(positive_for_drug)),
            positive_tests = sum(positive_for_drug, na.rm = TRUE),
            percent_positive = round(100 * positive_tests/n_not_NA,1))

plot_f2 <- fig2 %>% 
  ggplot(aes(x = year, y = percent_positive, position = drug_type))+
  geom_point(aes(shape = drug_type))+
  geom_line()+
  theme_few()+
  scale_y_continuous(breaks = c(0, 4, 8, 12))+
  labs(x = "Year", y = "Positive for Drugs, %", shape = "Drug type")
plot_f2
```

###Fig. 3: Prevalence of cannabinoid drugs in driver fatalities by year and age group

```{r figure3, fig.width = 8}
fig3 <- clean_fars %>%
  filter(drug_type == "Cannabinoid") %>% 
  filter(!is.na(agecat)) %>% 
  group_by(agecat, year) %>% 
  summarize(n_not_NA = sum(!is.na(positive_for_drug)),
            positive_tests = sum(positive_for_drug, na.rm = TRUE),
            percent_positive = round(100 * positive_tests/n_not_NA,1))

plot_f3 <- fig3 %>% 
  ggplot(aes(x = year, y = percent_positive, position = agecat))+
  geom_line()+
  geom_point(aes(shape = agecat))+
  theme_few()+
  labs(x = "Year", y = "Positive for Cannabinoid, %", shape = "Age")
plot_f3
```

###Prevalence of drugs in driver fatalities for 1999 and 2010 by drug type

```{r table_2, fig.width = 8}
datacis <- function(x, n) {
  p <- x / n
  standard_error <- sqrt((p * (1 - p)) / n)
  upper_ci <- (p + (1.96 * standard_error))
  lower_ci <- (p - (1.96 * standard_error))
  ptrans <- round((p * 100), digits = 1)
  upper_ci_trans <- round((upper_ci * 100), digits = 1)
  lower_ci_trans <- round((lower_ci * 100), digits = 1)
  results <- paste0(ptrans, "% (", lower_ci_trans, 
                          "%, ", upper_ci_trans, "%)")
  return(results)
}

t1 <- clean_fars %>% 
  filter(year %in% c("1999","2010")) %>% 
  group_by(year, drug_type) %>% 
  summarize(positive = sum(positive_for_drug, na.rm = TRUE), trials = sum(!is.na(positive_for_drug)))

t2 <- data.frame(prevalence = datacis(x = t1$positive, n = t1$trials))

t1 <- rowid_to_column(as.data.frame(t1, var = "rowid"))
t2 <- rowid_to_column(as.data.frame(t2, var = "rowid"))

t3 <- t1 %>% 
  full_join(t2, by = "rowid") %>% 
  select(prevalence, drug_type, year) %>% 
  spread(key = year, value = prevalence) %>% 
  rename("Drug type" = drug_type)

kable(t3) 
```

##Statistics for trends in drug prevalence by drug type using Cochran-Armitage trend test

```{r table_3, fig.width = 8}
ca_test <- function(drug, data = clean_fars) {
  if(drug == "Nonalcohol"){
    testing <- clean_fars %>%
      filter(drug_type != "Alcohol") %>%
      group_by(unique_id, year) %>% 
      summarize(positive_test = sum(positive_for_drug, na.rm = TRUE), 
                positive = any(positive_test > 0),
                total_tests = length(!is.na(positive_for_drug))) %>% 
      ungroup() %>% 
      group_by(year) %>%
      summarize(total_tests = sum(total_tests), positive = sum(positive)) 
  } else{
    testing <- clean_fars %>%
      filter(drug_type == drug) %>%
      group_by(year) %>%
      summarize(positive = sum(positive_for_drug, na.rm = TRUE),
                total_tests = sum(!is.na(positive_for_drug)))
  }
  ca_oh <- prop.trend.test(x = testing$positive,
                                n = testing$total_tests)
  Z <- round(sqrt(ca_oh$statistic), digits = 1)
  p.value <- round(ca_oh$p.value, digits = 3)
  ca_results <- data.frame(Z, p.value)
  tibble::remove_rownames(ca_results)
  return(ca_results)
}

dlist <- c("Alcohol", "Nonalcohol", "Narcotic", "Depressant",
               "Stimulant", "Cannabinoid", "Other")
drug_ca <- lapply(dlist, ca_test)
drug_ca <- dplyr::bind_rows(drug_ca) %>%
  dplyr::mutate(drug = dlist) %>%
  dplyr::select(drug, Z, p.value)
drug_ca %>% knitr::kable()
```

###Statistics for trends in drug prevalence by drug type using Wald test of logistic regression coefficient for "year"

```{r table_4, fig.width = 8}
test_reg <- function(drug, data = clean_fars) {
  if(drug == "Nonalcohol"){
    testing <- clean_fars %>%
      filter(!is.na(drug_type)) %>% 
      filter(drug_type != "Alcohol") %>% 
      group_by(unique_id, year) %>% 
      summarize(positive_for_drug = any(positive_for_drug))
    
  } else{
    testing <- clean_fars %>%
      filter(!is.na(drug_type)) %>%
      filter(drug_type == drug)
  }
  reg <- glm(positive_for_drug ~ year, data = testing,
                 family = binomial(link = "logit"))
  sum <- slice(broom::tidy(reg), 2)
  Z <- round(sum$statistic, digits = 1)
  p.value <- round(sum$p.value, digits = 3)
  results <- data.frame(Z, p.value)
  tibble::remove_rownames(results) 
  return(results)
}

dlist <- c("Alcohol", "Nonalcohol", "Narcotic", "Depressant",
"Stimulant", "Cannabinoid", "Other")
drug_wald <- lapply(dlist, test_reg)
drug_wald <- dplyr::bind_rows(drug_wald) %>%
dplyr::mutate(drug = dlist) %>%
dplyr::select(drug, Z, p.value)
drug_wald %>% knitr::kable()
```
